---
title: "Problem Set 2"
author: "Romain Fernex"
date: "2025-02-14"
output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex
    toc: true
  word_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
header-includes: 
  - \usepackage{titlesec}
  - \usepackage{fontspec}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

# Problem 1 


```{r setup, include=FALSE}

options(repos = c(CRAN = "https://cran.rstudio.com/")) # necessary to load the stargazer package
knitr::opts_chunk$set(comment = "") # improves output visuals by removing ##
# knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

if (!require(formatR))install.packages("formatR", repos="https://cran.rstudio.com/", dependencies = FALSE)
if (!require(haven)) install.packages("haven")
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(labelled)) install.packages("labelled")
if (!require(tidyr)) install.packages("tidyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(skimr)) install.packages("skimr")
if (!require(knitr)) install.packages("knitr")
if (!require(stargazer)) install.packages("stargazer")
if (!require(KableExtra)) install.packages("KableExtra")
if (!require(rmarkdown)) install.packages("rmarkdown")
library(formatR)
library(kableExtra)
library(haven)
library(tidyverse)
library(labelled)
library(tidyr)
library(ggplot2)
library(dplyr)
library(skimr)
library(knitr)
library(stargazer)
```

## Load the data
```{r, include=TRUE}
load("/Users/rfernex/Documents/Education/SciencesPo/Courses/S2/Econometrics/TD/Psets/Pset2/Data/401k.Rdata")
```

## Question 1 
```{r, include=TRUE}
mean_participation <- mean(data$prate)
mean_match <- mean(data$mrate)
cat("average participation rate :", mean_participation)
cat("average match rate :", mean_match)
```

## Question 2 
```{r, include=TRUE}
model <- lm(prate ~ mrate, data = data)

stargazer(model, type = "text",
          title = "Regression Summary: prate vs mrate",
          dep.var.labels = "Participation rate",
          covariate.labels = c("Match rate", "Intercept"),
          omit.stat = c("f"))

num_obs <- nobs(model)
r_squared <- summary(model)$r.squared

cat("Sample Size :", num_obs)
cat("R-squared :", sprintf("%.2f%%", r_squared*100))
```

## Question 3 
```{r, include=TRUE}
data$fitted_values <- fitted(model)
sum_of_residuals <- sum(residuals(model))
cat("Sum of Residuals:", sprintf("%.2f", sum_of_residuals), "\
")
kable(head(data$fitted_values, 10), col.names = c("Fitted Values"))
```


## Question 4 
```{r, include=TRUE}
intercept <- coef(model)["(Intercept)"]
coefficient_mrate <- coef(model)["mrate"]

cat("Intercept :", sprintf("%.2f", intercept), "\
")
cat("Coefficient for the match rate :", sprintf("%.2f", coefficient_mrate), "\
")
```
Based on the coefficient for the match rate, increasing the match rate by 1 percent raises the participation rate by 5.86 percent. 
The intercept is at 83.08%, this means that the predicted participation rate will always be superior or equal to 83%. This seems to indicate that participation rate of eligible workers tends to be pretty high regardless of their company's generosity. 
 
## Question 5 
```{r, include=TRUE}
source_mrate = 3.5 # adaptable
predicted_y <- intercept + coefficient_mrate * source_mrate
cat("Predicted participation rate for a match rate of", source_mrate, ":", predicted_y)
```
The predicted participation rate stands above a hundred which is unreasonable. 
\textbf{potential issues}
\begin{itemize}
\item The value used is outside the fitted data range which may lead to unreasonable predictions
\item It could also be possible that the relation between the regressor and the dependent variable is in fact not linear. (misspecification) 

\end{itemize} 
## Question 6
```{r, include=TRUE}
cat("The match rate explains", sprintf("%.2f", r_squared*100),"% of the variation in participation rate")
```
This seems like a rather small number given the fact the the match rate indicates how much the firm is willing to contribute to the pension plan. This means more than 90% of the variation in prate is attributable to other factors. 
Nonetheless, it does not mean that the match rate is not significantly related to the participation rate. 

## Question 7 
```{r, include=TRUE}
ggplot(data, aes(x = mrate, y = prate)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(
    title = "Scatter of prate as a function of mrate",
    x = "Match Rate (mrate)",
    y = "Participation Rate (prate)"
  ) +
  theme_minimal()
```
The Scatter plot seems to suggest a positive, non-linear relation between the participation rate and the match rate; 

```{r, include=TRUE}
# Creates column with binned values
data <- data %>% mutate(rnd_mrate = round(mrate, digits = 1),
                        mrate_binned = cut(rnd_mrate, 
                                           breaks = seq(min(rnd_mrate, na.rm = TRUE), 
                                                        max(rnd_mrate, na.rm = TRUE), 
                                                        by = 0.1), 
                                           include.lowest = TRUE))

# Calculate the average prate by bin
Average_mrate_df <- data %>% 
  group_by(mrate_binned) %>% 
  summarise(avg_prate_by_bin = mean(prate, na.rm = TRUE))

# Add predicted values to the dataset
data <- data %>% mutate(predicted_values = predict(model))

# Calculate the average predicted prate by bin
Predicted_mrate_df <- data %>% 
  group_by(mrate_binned) %>% 
  summarise(predicted_prate_by_bin = mean(predicted_values, na.rm = TRUE))

# Merge the two dataframes for plotting
plot_data <- Average_mrate_df %>%
  left_join(Predicted_mrate_df, by = "mrate_binned")

# Create the scatterplot with average prate and predicted prate
ggplot() +
  # Scatterplot for average prate
  geom_point(data = Average_mrate_df, aes(x = as.numeric(mrate_binned), y = avg_prate_by_bin), color = "blue") +
  # Line for predicted prate
  geom_line(data = Predicted_mrate_df, aes(x = as.numeric(mrate_binned), y = predicted_prate_by_bin), color = "red") +
  theme_minimal() +
  labs(title = "Scatter plot of average participation rate and predicted participation rate by match rate",
       x = "Match rate (binned)",
       y = "Participation rate") +
  scale_x_continuous(breaks = seq(0, max(as.numeric(Average_mrate_df$mrate_binned), na.rm = TRUE), by = 5)) +
  theme(panel.grid.major = element_line(color = "gray", linewidth = 0.5),
        panel.grid.minor = element_blank())
```
We notice clearly the positive relation between match rate and participation rate. However, based on the scatterplot, this relation is not linear and there remains a few outliers. 
Thus we can establish that the problem that our model suffers from is a  misspecification problem. 
